# Adverse Impact of Severe Weather Events on Public Health and Economy Between 1950 and 2011


```{r load}
data <- read.csv("C:\\Users\\Razib\\Desktop\\Reproducible Research\\2\\StormData\\StormData.csv", sep=",",quote="\"", na.strings = "NA")
dim(data)
head(data)
str(data)
colnames(data)
## Subset data
subData <- data[, c("STATE", "EVTYPE", "FATALITIES", "INJURIES", 
               "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]
head(subData)
```

Remove inconsequential values
```{r expldata}
## Check to see if there are rows that have value for all four columns
## FATALITIES and INJURIES and PROPDMG and CROPDMG is zero
ignore <- subData[which(subData$FATALITIES == 0 & subData$INJURIES == 0 &
                         subData$PROPDMG == 0 & subData$CROPDMG == 0), ]
## Ignore these rows that have values for all four columns to be zero because
## for the purpose of this analysis, these rows are inconsequential for their
## corresponding event types as they did not lead to any damage health or
## property.  So, select all rows that have non-zero value for at least one of
## these four columns.
myData <- subData[which(subData$FATALITIES > 0 | subData$INJURIES > 0 |
                         subData$PROPDMG > 0 | subData$CROPDMG > 0), ]
```

Clean event types according to specification in document.

Here we are looking to clean events that can be categrorized as TORNADO.
```{r checkcleantornado}
tornado <- myData[grepl("torn", myData$EVTYPE, ignore.case = TRUE),]
#Check to see if there are any event types to be cleaned
chktornado <- myData[grepl("torn", myData$EVTYPE, ignore.case = TRUE) & 
                             !grepl("tornado", 
                                    myData$EVTYPE, ignore.case = TRUE),]
## There is one event type where TORNADO  is misspelled as TORNDAO.  Clean in
## by substituting with correct spelling
myData[grepl("TORNDAO", 
             myData$EVTYPE, ignore.case = TRUE), "EVTYPE"] <- "TORNADO"
myData[grepl("TORNTORNDAO", 
             myData$EVTYPE, ignore.case = TRUE), "EVTYPE"] <- "TORNADO"
## Clean observations as:
## TORNADOES, TSTM WIND, HAIL -> TORNADO,
## TORNADOES -> TORNADO
myData[grepl("tornadoes", 
             myData$EVTYPE, ignore.case = TRUE), "EVTYPE"] <- "TORNADO"
myData[grepl("COLD AIR TORNADO", 
             myData$EVTYPE, ignore.case = TRUE), "EVTYPE"] <- "TORNADO"
myData[grepl("TORNADO F", 
             myData$EVTYPE, ignore.case = TRUE), "EVTYPE"] <- "TORNADO"
```

Categorize Avalance related events.
```{r checkcleanavlnch}
avl <- myData[grepl("aval", myData$EVTYPE, ignore.case = TRUE) &
                      !grepl("Avalanche", myData$EVTYPE, ignore.case = TRUE),]
## One entry where Avalanche is mispelled as Avalance.  Clean it up by 
## replacing with correct spelling.
myData$EVTYPE <- sub("AVALANCE", "AVALANCHE", myData$EVTYPE)
```

Categorize BLIZZARD related events.
```{r checkcleanblzrd, cache=TRUE}
## Explore blizzard related data
bl <- myData[grepl("bli", myData$EVTYPE, ignore.case = TRUE), ]
blz <- myData[grepl("bli", myData$EVTYPE, ignore.case = TRUE) &
                      !grepl("Blizzard", myData$EVTYPE, ignore.case = TRUE),]
myData[grepl("BLIZZARD/WINTER STORM", myData$EVTYPE, 
             ignore.case = TRUE), "EVTYPE"] <- "BLIZZARD"
myData[grepl("GROUND BLIZZARD", myData$EVTYPE, 
             ignore.case = TRUE), "EVTYPE"] <- "BLIZZARD"
## All rows for blizzard are already correct.  No cleaning needed.
```

Check and clean COASTAL FLOOD related events.
```{r chkclncoastalflood, cache=TRUE}
cf <- myData[grepl("^coastal flood", myData$EVTYPE, ignore.case = TRUE), ]
## All rows for Coastal flood are spelled correct.  But there are some rows
## that contain Coastal flooding and some rows have coastal flood/erosion instead
## of Coastal Flood that is specified in the in documentaion.  Clean it so that
## all these event types comply to the accompanying code book and so have value
## Coastal flood
myData$EVTYPE <- sub("coastal flooding", "Coastal Flood", myData$EVTYPE,
                     ignore.case = TRUE)
myData$EVTYPE <- sub("coastal flooding/erosion", "Coastal Flood", myData$EVTYPE,
                     ignore.case = TRUE)
myData$EVTYPE <- sub("coastal flood/erosion", "Coastal Flood", myData$EVTYPE,
                     ignore.case = TRUE)
```

Work on events related to COLD/WIND CHILL.
```{r chkclncoldwindchill}
cwc <- myData[grepl("^cold/", myData$EVTYPE, ignore.case = TRUE), ]
myData$EVTYPE <- sub("^cold/winds", "COLD/WIND CHILL", myData$EVTYPE, 
                     ignore.case = TRUE)
uev <- unique(myData$EVTYPE)
ucld <- uev[grepl("cold", uev, ignore.case = TRUE)]
myData[grepl("cold", myData$EVTYPE, 
                    ignore.case = TRUE) &
               !grepl("COLD/WIND CHILL", myData$EVTYPE, 
                    ignore.case = TRUE) &
               !grepl("EXTREME COLD/WIND CHILL", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "COLD/WIND CHILL"
```
Check and clean events related to DEBRIS FLOW.
```{r chkclndebrisflow}
deb <- myData[grepl("^deb", myData$EVTYPE, ignore.case = TRUE), ]
## No rows for Debris flow
```
Work on events related to DENSE FOG.
```{r chkclndnsfg}
dfg <- myData[grepl("^dens", myData$EVTYPE, ignore.case = TRUE), ]
chkdfg <- myData[grepl("^dens", myData$EVTYPE, ignore.case = TRUE) &
                         !grepl("^dense fog", myData$EVTYPE, 
                                ignore.case = TRUE), ]

```
Work on evens related to DROUGHT.
```{r chkclndrought}
drt <- myData[grepl("drou", myData$EVTYPE, ignore.case = TRUE), ]
myData$EVTYPE <- sub("DROUGHT/EXCESSIVE HEAT", "DROUGHT", myData$EVTYPE, 
                     ignore.case = TRUE)
drtcln <- myData[grepl("drou", myData$EVTYPE, ignore.case = TRUE) &
                         !grepl("^drou", myData$EVTYPE, ignore.case = TRUE), ]
myData$EVTYPE <- sub("HEAT WAVE DROUGHT", "DROUGHT", myData$EVTYPE, 
                     ignore.case = TRUE)
````
Work on events related to DUST DEVIL.
```{r chkclndstdvl}
dd <- myData[grepl("dust devil", myData$EVTYPE, ignore.case = TRUE), ]
myData$EVTYPE <- sub("DUST DEVIL WATERSPOUT", "DUST DEVIL", 
                     myData$EVTYPE, ignore.case = TRUE)
ddchk <- myData[grepl("dust devil", myData$EVTYPE, ignore.case = TRUE) &
                     !grepl("^dust", myData$EVTYPE, ignore.case = TRUE) , ]
```
Work on events related to DUST STORM.
```{r chkclndststol}

ds <- myData[grepl("dust storm", myData$EVTYPE, ignore.case = TRUE), ]
dschk <- myData[grepl("dust", myData$EVTYPE, ignore.case = TRUE) &
                        !grepl("^dust", myData$EVTYPE, ignore.case = TRUE), ]
myData$EVTYPE <- sub("BLOWING DUST", "DUST STORM", 
                     myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("DUST STORM/HIGH WINDS", "DUST STORM", 
                     myData$EVTYPE, ignore.case = TRUE)
```
Work on events related to EXCESSIVE HEAT.
```{r chkclnexht}
eh <- myData[grepl("excessive heat", myData$EVTYPE, ignore.case = TRUE), ]

ehchk <- myData[grepl("excessive heat", myData$EVTYPE, ignore.case = TRUE) &
                         !grepl("^excessive heat", myData$EVTYPE, 
                                ignore.case = TRUE), ]
myData$EVTYPE <- sub("RECORD/EXCESSIVE HEAT", "EXCESSIVE HEAT", 
                     myData$EVTYPE, ignore.case = TRUE)
ehp <- myData[grepl("excessive heat", myData$EVTYPE, ignore.case = TRUE) &
                      !grepl("^excessive heat", myData$EVTYPE, 
                                ignore.case = TRUE), ]
```
Work on events related to EXTREME COLD/WIND CHILL.
```{r chkclnextcld}
ec <- myData[grepl("extr", myData$EVTYPE, ignore.case = TRUE), ]

echks <- myData[grepl("extre", myData$EVTYPE, ignore.case = TRUE) &
                         !grepl("^extre", myData$EVTYPE, ignore.case = TRUE), ]

echk <- myData[grepl("extre", myData$EVTYPE, ignore.case = TRUE) &
                         !grepl("extreme cold", myData$EVTYPE, 
                                ignore.case = TRUE), ]
myData$EVTYPE <- sub("EXTREME WIND CHILL", "EXTREME COLD/WIND CHILL", 
                     myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("EXTREME WINDCHILL", "EXTREME COLD/WIND CHILL", 
                     myData$EVTYPE, ignore.case = TRUE)
myData[grepl("EXTREME COLD", myData$EVTYPE, ignore.case = TRUE),
       "EVTYPE"] <- "EXTREME COLD/WIND CHILL"
```
Work on events related FLASH FLOOD.
```{r chkclnff}
fl <- myData[grepl("flash", myData$EVTYPE, ignore.case = TRUE), ]
myData$EVTYPE <- sub("FLASH FLOODING", "FLASH FLOOD", 
                     myData$EVTYPE, ignore.case = TRUE)
ffp <- myData[grepl("flash", myData$EVTYPE, ignore.case = TRUE), ]
ffchk <- myData[grepl("flash", myData$EVTYPE, ignore.case = TRUE) &
                         !grepl("^flash", myData$EVTYPE, 
                                ignore.case = TRUE), ]
uev <- unique(myData$EVTYPE)
flfl <- uev[grepl("flash", uev, ignore.case = TRUE)]
myData[grepl("ICE STORM/FLASH FLOOD", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "ICE STORM"
myData[grepl("FLASH FLOOD WINDS", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "FLASH FLOOD"
myData[grepl("FLASH FLOOD/", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "FLASH FLOOD"
myData[grepl("FLASH FLOODS", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "FLASH FLOOD"
myData[grepl("URBAN AND FLASH FLOODIN", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "FLASH FLOOD"
myData[grepl("MUD SLIDES FLASH FLOOD", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "FLASH FLOOD"
myData[grepl("FLASH FLOOD FROM ICE JAMS", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "FLASH FLOOD"
myData[grepl("FLASH FLOOD - HEAVY RAIN", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "FLASH FLOOD"
myData[grepl("FLOOD FLASH", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "FLASH FLOOD"
myData[grepl("FLOOD/FLASH", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "FLASH FLOOD"
myData[grepl("FLASH FLOOD LANDSLIDES", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "FLASH FLOOD"
myData[grepl("FLASH FLOOD \\(MINOR", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "FLASH FLOOD"
myData[grepl(" FLASH FLOOD", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "FLASH FLOOD"
myData[grepl("URBAN AND FLASH FLOODIN", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "FLASH FLOOD"

```
Work on events related to FLOOD.
```{r chkclnfld}
fld <- myData[grepl("^flood", myData$EVTYPE, ignore.case = TRUE), ]
myData$EVTYPE <- sub("FLOODING", "FLOOD", myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("FLOOD/FLASH FLOOD", "FLOOD", 
                     myData$EVTYPE, ignore.case = TRUE)
fldp <- myData[grepl("flood", myData$EVTYPE, ignore.case = TRUE) &
                       !grepl("^flood", myData$EVTYPE, ignore.case = TRUE) &
                       !grepl("flash flood", myData$EVTYPE, ignore.case = TRUE) &
                       !grepl("coastal flood", myData$EVTYPE, 
                              ignore.case = TRUE) &
                       !grepl("lakeshore flood", myData$EVTYPE, 
                              ignore.case = TRUE), ]
myData$EVTYPE <- sub("RIVER FLOOD", "FLOOD", myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("BREAKUP FLOOD", "FLOOD", myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("URBAN FLOOD", "FLASH FLOOD", myData$EVTYPE, 
                     ignore.case = TRUE)
myData$EVTYPE <- sub("URBAN/SMALL STREAM FLOOD", "FLASH FLOOD", myData$EVTYPE, 
                     ignore.case = TRUE)
myData$EVTYPE <- sub("MAJOR FLOOD", "FLOOD", myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("MINOR FLOOD", "FLOOD", myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("RURAL FLOOD", "FLOOD", myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("ICE JAM FLOOD", "FLASH FLOOD", myData$EVTYPE, 
                     ignore.case = TRUE)
myData$EVTYPE <- sub("SMALL STREAM FLOOD", "FLASH FLOOD", myData$EVTYPE, 
                     ignore.case = TRUE)
myData$EVTYPE <- sub("LAKE FLOOD", "LAKESHORE FLOOD", myData$EVTYPE, 
                     ignore.case = TRUE)
myData$EVTYPE <- sub("URBAN AND SMALL STREAM FLOODIN", "FLASH FLOOD", 
                     myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("RIVER AND STREAM FLOOD", "FLASH FLOOD", 
                     myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("SNOWMELT FLOOD", "FLASH FLOOD", 
                     myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("HEAVY SNOW/HIGH WINDS & FLOOD", "HEAVY SNOW", 
                     myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("HEAVY RAIN AND FLOOD", "HEAVY RAIN", 
                     myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("HEAVY RAINS/FLOOD", "HEAVY RAIN", 
                     myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("THUNDERSTORM WINDS/FLOOD", "THUNDERSTORM WIND", 
                     myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("Erosion/Cstl Flood", "COASTAL FLOOD", 
                     myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("COASTAL  FLOOD/EROSION", "COASTAL FLOOD", 
                     myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("TIDAL FLOOD", "FLOOD", 
                     myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("THUNDERSTORM WINDS/ FLOOD", "THUNDERSTORM WIND", 
                     myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("FLASH FLOOD/THUNDERSTORM WI", "FLASH FLOOD", 
                     myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("FLOODS", "FLOOD", 
                     myData$EVTYPE, ignore.case = TRUE)
myData$EVTYPE <- sub("FLOOD/FLOOD", "FLOOD", 
                     myData$EVTYPE, ignore.case = TRUE)

fldps <- myData[grepl("flood", myData$EVTYPE, ignore.case = TRUE) &
                       !grepl("^flood", myData$EVTYPE, ignore.case = TRUE) &
                       !grepl("flash flood", myData$EVTYPE, ignore.case = TRUE) &
                       !grepl("coastal flood", myData$EVTYPE, 
                              ignore.case = TRUE) &
                       !grepl("lakeshore flood", myData$EVTYPE, 
                              ignore.case = TRUE), ]
```
Check and clean events related to THUDERSTORM WIND.
```{r chkclntstm}
ts <- myData[grepl("^thun", myData$EVTYPE, ignore.case = TRUE), ]
myData[grepl("^thun", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "THUNDERSTORM WIND"
tschk <- myData[grepl("^thun", myData$EVTYPE, ignore.case = TRUE) &
                        grepl("thunderstorm winds", myData$EVTYPE, 
                              ignore.case = TRUE), ]

tsb <- myData[grepl("thun", myData$EVTYPE, ignore.case = TRUE) &
                     !grepl("^thun", myData$EVTYPE, ignore.case = TRUE), ]

myData[grepl("severe thun", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "THUNDERSTORM WIND"
tsbchk <- myData[grepl("thun", myData$EVTYPE, ignore.case = TRUE) &
                     !grepl("^thun", myData$EVTYPE, ignore.case = TRUE), ]
myData[grepl("LIGHTNING THUNDERSTORM WINDS", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "THUNDERSTORM WIND"
myData[grepl("LIGHTNING AND THUNDERSTORM WIN", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "THUNDERSTORM WIND"
myData[grepl("TUNDERSTORM WIND", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "THUNDERSTORM WIND"
myData[grepl("THUDERSTORM WINDS", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "THUNDERSTORM WIND"
myData[grepl("^tstm", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "THUNDERSTORM WIND"
tstm <- myData[grepl("TSTM", myData$EVTYPE, ignore.case = TRUE), ]
myData[grepl("MARINE TSTM WIND", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "MARINE THUNDERSTORM WIND"
myData[grepl("NON-TSTM WIND", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "HIGH WIND"
myData[grepl("TSTM WIND", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "THUNDERSTORM WIND"
tstmchk <- myData[grepl("TSTM", myData$EVTYPE, ignore.case = TRUE), ]
```
Work on events related to FROST/FREEZE.
```{r chkclnfrst}
uev <- unique(myData$EVTYPE)
frst <- uev[grepl("frost", uev, ignore.case = TRUE)]

myData[grepl("frost", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "FROST/FREEZE"
myData[grepl("freeze", myData$EVTYPE, 
                    ignore.case = TRUE) &
               !grepl("FROST/FREEZE", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "FROST/FREEZE"
```
Work on events related to HAIL.
```{r chkclnhl}
uev <- unique(myData$EVTYPE)
hl <- uev[grepl("hail", uev, ignore.case = TRUE)]
myData[grepl("^hail", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "HAIL"
myData[grepl("WIND/HAIL", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "HAIL"
myData[grepl("SMALL HAIL", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "HAIL"
myData[grepl("GUSTY WIND/HAIL", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "HAIL"
```
Work on events related to HEAT.
```{r chkclnht}
uev <- unique(myData$EVTYPE)
ht <- uev[grepl("heat", uev, ignore.case = TRUE)]
myData[grepl("heat", myData$EVTYPE, 
                    ignore.case = TRUE) &
               !grepl("excessive heat", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "HEAT"
```
Work on events related to HEAVY RAIN.
```{r chkclnhvrn}
uev <- unique(myData$EVTYPE)
hvrn <- uev[grepl("rain", uev, ignore.case = TRUE)]
myData[grepl("rain", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "HEAVY RAIN"
```
Work on events related to HEAVY SNOW.
```{r chkclnhvsn}
uev <- unique(myData$EVTYPE)
hvsn <- uev[grepl("snow", uev, ignore.case = TRUE)]

myData[grepl("snow", myData$EVTYPE, 
                    ignore.case = TRUE) &
               !grepl("lake", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "HEAVY SNOW"
```
Work on events related to LAKE-EFFECT SNOW.
```{r chkclnlksn}
uev <- unique(myData$EVTYPE)
lksn <- uev[grepl("lake", uev, ignore.case = TRUE)]
## HEAVY LAKE SNOW -> LAKE-EFFECT SNOW,  Lake Effect Snow -> LAKE-EFFECT SNOW 
myData[grepl("HEAVY LAKE SNOW", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "LAKE-EFFECT SNOW"
myData[grepl("Lake Effect Snow", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "LAKE-EFFECT SNOW"
```
Work on events related to HIGH SURF.
```{r chkclnhsrf}
uev <- unique(myData$EVTYPE)
hsrf <- uev[grepl("tide", uev, ignore.case = TRUE)]
myData[grepl("surf", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "HIGH SURF"
```
Work on events related to HIGH WIND.
```{r chkclnhwnd}
uev <- unique(myData$EVTYPE)
wnd <- uev[grepl("wind", uev, ignore.case = TRUE)]
uev <- unique(myData$EVTYPE)
clnwnd <- uev[grepl("wind", uev, ignore.case = TRUE) &
                   !grepl("HURRICANE OPAL", uev, ignore.case = TRUE) &
                   !grepl("MARINE", uev, ignore.case = TRUE) &
                   !grepl("WINTER STORM", uev, ignore.case = TRUE) &
                   !grepl("Cold/Wind Chill", uev, ignore.case = TRUE) &
                   !grepl("Extreme Cold/Wind Chill", uev, ignore.case = TRUE) &
                   !grepl("Strong wind", uev, ignore.case = TRUE) &
                   !grepl("Thunderstorm Wind", uev, ignore.case = TRUE) &
                   !grepl("gusty wind", uev, ignore.case = TRUE)]
myData[grepl("wind", myData$EVTYPE, 
                    ignore.case = TRUE) &
               !grepl("HURRICANE OPAL", myData$EVTYPE, 
                    ignore.case = TRUE) &
               !grepl("MARINE", myData$EVTYPE, 
                    ignore.case = TRUE) &
               !grepl("WINTER STORM", myData$EVTYPE, 
                    ignore.case = TRUE) &
               !grepl("Cold/Wind Chill", myData$EVTYPE, 
                      ignore.case = TRUE) &
               !grepl("Extreme Cold/Wind Chill", myData$EVTYPE, 
                      ignore.case = TRUE) &
               !grepl("Strong wind", myData$EVTYPE, 
                      ignore.case = TRUE) &
               !grepl("Thunderstorm Wind", myData$EVTYPE, 
                      ignore.case = TRUE) &
               !grepl("gusty wind", myData$EVTYPE, 
                      ignore.case = TRUE), "EVTYPE"] <- "HIGH WIND"

```
Work on events related to HURRICANE (TYPHOON)
```{r chkclnhrcn}
uev <- unique(myData$EVTYPE)
hrcn <- uev[grepl("hur", uev, ignore.case = TRUE)]
myData[grepl("hur", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "HURRICANE (TYPHOON)"
myData[grepl("TYPHOON", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "HURRICANE (TYPHOON)"
```
Work on events related to ICE STORM.
```{r chkclnicst}
uev <- unique(myData$EVTYPE)
icst <- uev[grepl("ice", uev, ignore.case = TRUE)]
myData[grepl("ice", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "ICE STORM"
```
Work on events related to LIGHTNING.
```{r chkclnltng}
uev <- unique(myData$EVTYPE)
ltng <- uev[grepl("light", uev, ignore.case = TRUE)]
myData[grepl("light", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "LIGHTNING"
myData[grepl("LIGNTNING", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "LIGHTNING"
```
Work on events related to MARINE HIGH WIND.
```{r chkclnmrn}
uev <- unique(myData$EVTYPE)
mrn <- uev[grepl("marine", uev, ignore.case = TRUE)]
myData[grepl("Marine Accident", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "MARINE HIGH WIND"
myData[grepl("MARINE MISHAP", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "MARINE HIGH WIND"
```
Work on events related to RIP CURRENT.
```{r chkclnrpcr}
uev <- unique(myData$EVTYPE)
rpcr <- uev[grepl("rip", uev, ignore.case = TRUE)]
## Values RIP CURRENTS of EVTYPE to be changed to RIP CURRENT.  
myData[grepl("RIP CURRENTS", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "RIP CURRENT"
```
Work on events related to STORM SURGE/TIDE.
```{r chkclnstsg}
uev <- unique(myData$EVTYPE)
stsg <- uev[grepl("Surge", uev, ignore.case = TRUE)]
## Two Values STORM SURGE and "COASTAL SURGE of EVTYPE need to be made compliant  
myData[grepl("STORM SURGE", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "STORM SURGE/TIDE"
myData[grepl("COASTAL SURGE", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "STORM SURGE/TIDE"
```
Work on events related to STRONG WIND.
```{r chkclnstwnd}
uev <- unique(myData$EVTYPE)
stwnd <- uev[grepl("Strong", uev, ignore.case = TRUE)]

myData[grepl("Strong winds", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "STRONG WIND"
uev <- unique(myData$EVTYPE)
gswnd <- uev[grepl("gusty", uev, ignore.case = TRUE)]

myData[grepl("gusty", myData$EVTYPE, 
                    ignore.case = TRUE), "EVTYPE"] <- "STRONG WIND"
```
Work on events related to TROPICAL STORM.
```{r chkclntrst}
uev <- unique(myData$EVTYPE)
trst <- uev[grepl("trop", uev, ignore.case = TRUE)]


myData[grepl("trop", myData$EVTYPE, ignore.case = TRUE) &
               !grepl("tropical depression", myData$EVTYPE, 
                      ignore.case = TRUE), "EVTYPE"] <- "TROPICAL STORM"
```
Work on events related to WATERSPOUT.
```{r chkclnws}
uev <- unique(myData$EVTYPE)
ws <- uev[grepl("water", uev, ignore.case = TRUE)]


myData[grepl("water", myData$EVTYPE, 
             ignore.case = TRUE), "EVTYPE"] <- "WATERSPOUT"
```
Work on events related to WILDFIRE.
```{r chkclnwf}
uev <- unique(myData$EVTYPE)
wf <- uev[grepl("wild", uev, ignore.case = TRUE)]


myData[grepl("wild", myData$EVTYPE, 
             ignore.case = TRUE), "EVTYPE"] <- "WILDFIRE"
```
Work on events related to WINTER WEATHER.
```{r chkclnwth}
uev <- unique(myData$EVTYPE)
wth <- uev[grepl("weather", uev, ignore.case = TRUE)]
myData[grepl("WARM WEATHER", myData$EVTYPE, 
             ignore.case = TRUE), "EVTYPE"] <- "HEAT"
myData[grepl("WEATHER", myData$EVTYPE, 
             ignore.case = TRUE), "EVTYPE"] <- "WINTER WEATHER"
```
Work on events related to WINTER STORM.
```{r chkclnwst}
uev <- unique(myData$EVTYPE)
wtst <- uev[grepl("winter", uev, ignore.case = TRUE)]
myData[grepl("winter", myData$EVTYPE, 
             ignore.case = TRUE) &
               !grepl("winter weather", myData$EVTYPE, 
                      ignore.case = TRUE), "EVTYPE"] <- "WINTER STORM"
```
Rest of the values were taken care of while working on events specified above.

Convert all events to be in upper case for consistency.
```{r ucase}
myData$EVTYPE <- toupper(myData$EVTYPE)
```

## Results

``` {r health}

library(ggplot2)

inj <- myData[,c("EVTYPE", "INJURIES")]

eventinj <- aggregate(INJURIES ~ EVTYPE, data=inj, FUN=sum)

eventinj <- eventinj[order(eventinj$INJURIES, decreasing = TRUE),]

eventinj <- eventinj[1:15, ]

ftlts <- myData[,c("EVTYPE", "FATALITIES")]

eventftl <- aggregate(FATALITIES ~ EVTYPE, data=ftlts, FUN=sum)
eventftl <- eventftl[order(eventftl$FATALITIES, decreasing = TRUE),]

eventftl <- eventftl[1:15, ]

injvec <- vector(mode="character", length=15)
injvec[1:15] <- "Injuries"
treventinj <- data.frame("Events"=eventinj$EVTYPE, 
                         "Number" = eventinj$INJURIES, 
                         "Consequence" = injvec)
ftlvec <- vector(mode="character", length=15)
ftlvec[1:15] <- "Fatalities"
treventftl <- data.frame("Events"=eventftl$EVTYPE, 
                         "Number" = eventftl$FATALITIES, 
                         "Consequence" = ftlvec)
combevent <- rbind(treventinj, treventftl)
combevent
ggplot(data=combevent, aes(x=Events, y=Number, fill=Events)) + 
        geom_bar(stat="identity") +
        xlab("Weather event") + ylab("Injuries/Fatalities per weather event") +
        ggtitle("Injuries and Fatalities by weather event in US") +
        theme(axis.text.x  = element_text(angle=90, vjust=0.5, hjust=1)) +
        facet_grid(Consequence ~ .)

suminft <- rowSums(cbind(myData$INJURIES, myData$FATALITIES))
tothlth <- data.frame("EVTYPE" = myData$EVTYPE, 
                 "HEALTH" = suminft)

eventsm <- aggregate(HEALTH ~ EVTYPE, data=tothlth, FUN=sum)

eventsm <- eventsm[order(eventsm$HEALTH, decreasing = TRUE),]

eventsm <- eventsm[1:15, ]
eventsm

ggplot(data=eventsm, aes(x=EVTYPE, y=HEALTH, fill=EVTYPE)) + 
        geom_bar(stat="identity") +
        xlab("Weather event") + 
        ylab("Sum of Injuries and Fatalities per event") +
        ggtitle("Sum of Injuries and Fatalities by weather event across US") +
        theme(axis.text.x  = element_text(angle=90, vjust=0.5, hjust=1))
```


### Assess across the United Statess, which weather event caused most damage to economy implying the most damage caused in terms of both Property Damage as well as Crop Damage

```{r expleconomy}

eco <- myData[, c("EVTYPE", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]
unqc <- unique(myData$CROPDMGEXP)
unqp <- unique(myData$PROPDMGEXP)
```
```{r convexp}
#base vector
base <- c(0, 1, 2, 3, 4, 5, 6, 7, 8, 9, "b", "B", "m", "M", "k", "K", "h", "H")
#conversion vector
conv <- c(1e-09, 1e-08, 1e-07, 1e-06, 1e-05, 1e-04, 1e-03, 0.01, 0.1, 1, 1, 1,
        1e-03, 1e-03, 1e-06, 1e-06, 1e-07, 1e-07)
##Create a data frame representing a conversion table
convtab <- data.frame(base=base, conv=conv)
## Subset the dataframe to contain values from bs only in PROPDMGEXP and
## CROPDMGEXP
clneco <- subset(eco, eco$CROPDMGEXP %in% convtab$base | 
                         eco$PROPDMGEXP %in% convtab$base)
## Add column for conversion of CROPDMNEXP
clneco$CROPDMGEXPCONV <- sapply(clneco$CROPDMGEXP, function(x)
        if(x %in% convtab$base) convtab[base == x, "conv"] else 0)
## Add column for conversion of PROPDMNEXP
clneco$PROPDMGEXPCONV <- sapply(clneco$PROPDMGEXP, function(x)
        if(x %in% convtab$base) convtab[base == x, 2] else 0)
```
```{r caldmg}
clneco <- transform(clneco, TOTDMG = CROPDMG * CROPDMGEXPCONV + 
                  PROPDMG * PROPDMGEXPCONV)
```
```{r eventdmg}
## Subset dataframe to contain event and total damage only
dmg <- data.frame("Event" = clneco$EVTYPE, "Damage" = as.numeric(clneco$TOTDMG))
eventdmg <- aggregate(Damage ~ Event, data=dmg, FUN=sum)
eventdmg <- eventdmg[order(eventdmg$Damage, decreasing = TRUE),]
## Take 15 events that caused most Damage for this analysis
eventdmg <- eventdmg[1:15, ]
eventdmg
```

```{r dmgplot}
library(ggplot2)
ggplot(data=eventdmg, aes(x=Event, y=Damage, fill=Event)) + 
        geom_bar(stat="identity") +
        xlab("Weather event") + 
        ylab("Total damage in Billions per weather event") +
        ggtitle("Total damage in billions by weather event across US") +
        theme(axis.text.x  = element_text(angle=90, vjust=0.5, hjust=1))
```
